---
title: "Elements of programming"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
page-layout: full
bibliography: biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    smooth-scroll: true
    fig-dpi: 90
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

## [Homepage](index.html)

```{r}
#| warning: false
#| echo: false
#| include: false
#| message: false
#| purl: false

# Please ignore this chunk.
# A single purl() call: calling it twice in the same knit would clash on the
# chunk labels. The clean script is derived from the full one, so the two
# cannot go out of sync.

# 1. Full script, for personal use: code plus the text of the notes as
#    comments. Not linked from the website (and ignored by git).
knitr::purl("un_B.qmd", output = "code/un_B_full.R", documentation = 2)
full <- readLines("code/un_B_full.R")
# strip the Quarto markup that is only meaningful in the rendered page
full <- full[!grepl("^#' *:::", full)]
full <- gsub("\\[([^]]*)\\]\\{[^}]*\\}", "\\1", full)
squeeze <- function(x) x[!(x == "" & c(FALSE, head(x == "", -1)))]
writeLines(squeeze(full), "code/un_B_full.R")

# 2. Clean script, linked from the website: code only, obtained by dropping
#    the text (#') and the chunk headers (#+).
clean <- squeeze(full[!grepl("^#'|^#\\+", full)])
writeLines(clean, "code/un_B.R")
styler:::style_file("code/un_B.R")
```

## Unit B

-   Defining a new function
-   Control statements: `if`, `else`
-   `for` loops, `while` loops
-   An outline of the `*apply` functions
-   **R** packages

::: callout-note
#### Note

The exercises associated with this unit are available at
[this link](https://tommasorigon.github.io/EnvEcoStat/exe/exe_A.html).
:::

## Functions I

Besides being a statistical software, **R** is a [programming language]{.orange}.

Most **R** objects are functions. Indeed:

```{r}
class(sum) # Identifies the type of object
```

```{r}
class(log) # Second example
```

It is therefore possible to define [new functions]{.blue}, in addition to the ones
that already exist.

Once created, functions can be used exactly like all the others.

## Functions II

To create a new function, one uses the command `function`.

The [new function]{.orange} `cube` computes the cube of the number it receives as
input:

```{r}
cube <- function(x) {
  out <- x^3
  out
}
```

The [last object]{.blue} (in this case called `out`) is returned as the result:

```{r}
cube(4) # Computes the cube of the value 4
```

Alternatively, we can also use the command `return` to return the result:

```{r}
cube <- function(x) {
  out <- x^3
  return(out) # Makes it explicit that the value to be returned is out
}
cube(8) # Computes the cube of the value 8
```

## Functions III

It is possible to define new functions with [multiple arguments]{.orange}, also
choosing their [default values]{.blue}

```{r}
power <- function(x, p = 2) {
  out <- x^p
  out
}
```

The default value for the argument `p` is the square (`p = 2`), indeed:

```{r}
power(x = 4) # Computes the square of the value 4
```

```{r}
power(4) # Alternative syntax: the argument names need not be specified
```

However, we can select a different power as follows:

```{r}
power(x = 4, p = 3) # Computes the cube of the value 4
power(4, 3) # Alternative syntax: the argument names need not be specified
```

## Control statements I

The control statement `if` makes it possible to perform a given operation only
if a certain [condition]{.orange} is satisfied (`TRUE` or `FALSE`).

The (optional) control statement `else` makes it possible to perform an
[alternative operation]{.blue} if the previous condition is not satisfied.

For example, the following code prints on screen (function `print`) a different
sentence depending on whether the condition is true or false:

```{r}
condition <- pi^2 < 10 # Boolean value (in this case the condition is TRUE)

if (condition) {
  print("The condition is true")
  # Some commands to be executed
  # ...
} else {
  print("The condition is false")
  # Other commands to be executed
  # ...
}
```

## Control statements II

The following function `square_root` is a more concrete example concerning `if`
and `else`.

If the number provided as input is negative, a `NaN` is returned:

```{r}
square_root <- function(x) {
  if (x < 0) {
    # Warning message; it would actually be more appropriate to use "warning"
    print("The value of x must be positive")
    out <- NaN # Return Not A Number
  } else {
    out <- sqrt(x)
  }
  out
}

square_root(-2) # The condition x < 0 is satisfied
square_root(36) # The condition x < 0 is NOT satisfied
```

## `while` and `for` loops I

`for` loops and `while` loops repeat the operation contained between the curly
brackets `{ }` until a given condition is satisfied.

Let us start with a simple example concerning the `while` function:

```{r}
i <- 5 # We start with i = 5

while (i <= 25) { # Repeats the operation as long as i is smaller or equal to 25
  print(i) # Prints on screen the value of i
  i <- i + 5 # Increases the value of i

  # In real settings, here there are of course other operations to be executed
  # ...
}
```

## `while` and `for` loops II

::: callout-warning
Be careful not to create [endless loops]{.orange}!
:::

The following code requires a forced interruption of the **R** session or, in
the worst case, a restart of the computer:

```{r}
#| eval: false
# DO NOT RUN THE FOLLOWING CODE!
#
# The condition i <= 25 is always true because i is never updated
i <- 5
while (i <= 25) {
  print(i)

  # Other operations to be executed
  # ...
}
```

## `while` and `for` loops III

As an alternative to `while` loops, one can use the more explicit syntax of
`for` loops.

The `for` loop executes the content of the curly brackets considering, one at a
time, the values contained for instance in a vector:

```{r}
values <- seq(from = 5, to = 25, by = 5)
values

for (i in values) {
  print(i + 2) # Prints on screen the value of i + 2

  # Other operations to be executed
  # ...
}
```

## Summary exercise I

Compute the elements of a square matrix ${\bf D}$ of dimension $n \times n$,
whose elements are equal to $$
d_{ij} = (x_i - x_j)^2, \qquad  i,j \in \{1,\dots,n\},
$$ where ${\bf x} = (x_1,\dots,x_n)^\intercal$ is a generic vector in
$\mathbb{R}^n$.

In other words, define a function `distances(x)` that, starting from a generic
vector `x`, returns a matrix `D`.

::: callout-tip
## Outline of the solution

First create a matrix `D` whose elements are all equal to `0`, using the command
`matrix`.

Then use two "nested" `for` loops, that is one inside the other, to compute each
of the values `D[i, j]`.
:::

#### Solution of summary exercise I

```{r}
distances <- function(x) {
  n <- length(x) # Get the length of the vector x
  D <- matrix(0, nrow = n, ncol = n) # Creation of an empty matrix

  for (i in 1:n) {
    for (j in 1:n) {
      D[i, j] <- (x[i] - x[j])^2
    }
  }
  D # Value to be returned
}

x <- c(5, 2, 1, 24) # Example to check that it is correct
distances(x)
```

::: callout-note
## Note

There are many different (but correct) ways of implementing this function.
Incidentally, this solution is by no means the most efficient one.
:::

## Summary exercise II

`Fizzbuzz` is a simple programming exercise often used in job interviews to
check basic programming skills.

The task is the following: for all the numbers from $1$ to $100$ print on
screen:

-   the word `fizz` if the number is a multiple of $3$,

-   the word `buzz` if it is a multiple of $5$,

-   the word `fizzbuzz` if the number is a multiple of both $3$ and $5$,

-   the number itself otherwise.

::: callout-tip
## Hint

Use the function `print` and the remainder function `%%` of **R**. Moreover, to
make the code leaner it is possible to use the control statement called
`else if`.
:::

#### Solution of summary exercise II

The solution below makes use of the `else if` construct.

```{r}
#| eval: false
for (i in 1:100) {
  condA <- (i %% 3) == 0 # Is the number a multiple of 3?
  condB <- (i %% 5) == 0 # Is the number a multiple of 5?

  if (condA & condB) {
    print("fizzbuzz")
  } else if (condA) {
    print("fizz")
  } else if (condB) {
    print("buzz")
  } else {
    print(i)
  }
}
```

## The `*apply` family of functions

When possible, it would be better to [avoid]{.blue} using `for` loops, because they
tend to be slow in **R** (unlike other languages, such as C++).

A more elegant alternative, although not necessarily a more efficient one, is
the `*apply` family of functions, that is: `apply`, `tapply`, `sapply`,
`mapply`, `lapply`.

The simplest one, `apply`, executes a given function for each row / column of a
matrix.

::: callout-note
## Exercise

Consult the documentation of the `*apply` functions to get a first idea of how
they work.
:::

## **R** packages

**R** is organised in [packages]{.orange}.

If we want to use the functions of a package, it can be loaded using the
function `library`, that is:

```{r}
library(MASS) # Loads the MASS package into memory
library(knitr) # Loads the knitr package into memory
```

If a package is not available on the computer, it has to be installed.

One can use the following command, or use the "Packages" window available in
RStudio.

```{r}
#| eval: false
install.packages("knitr") # Installs the knitr package
```
