---
title: "Descriptive analysis of the dde data"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
page-layout: full
bibliography: biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    smooth-scroll: true
    fig-dpi: 90
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

## [Homepage](index.html)

```{r}
#| warning: false
#| echo: false
#| include: false
#| message: false
#| purl: false

# Please ignore this chunk.
# A single purl() call: calling it twice in the same knit would clash on the
# chunk labels. The clean script is derived from the full one, so the two
# cannot go out of sync.

# 1. Full script, for personal use: code plus the text of the notes as
#    comments. Not linked from the website (and ignored by git).
knitr::purl("un_D.qmd", output = "code/un_D_full.R", documentation = 2)
full <- readLines("code/un_D_full.R")
# strip the Quarto markup that is only meaningful in the rendered page
full <- full[!grepl("^#' *:::", full)]
full <- gsub("\\[([^]]*)\\]\\{[^}]*\\}", "\\1", full)
squeeze <- function(x) x[!(x == "" & c(FALSE, head(x == "", -1)))]
writeLines(squeeze(full), "code/un_D_full.R")

# 2. Clean script, linked from the website: code only, obtained by dropping
#    the text (#') and the chunk headers (#+).
clean <- squeeze(full[!grepl("^#'|^#\\+", full)])
writeLines(clean, "code/un_D.R")
styler:::style_file("code/un_D.R")
```

## Topics covered

-   Absolute and relative frequencies
-   The cumulative distribution function
-   Graphical representations (boxplots, histograms)
-   Location indices (mean, median, quantiles)

::: callout-note
#### Note

The exercises associated with this unit are available at
[this link](https://tommasorigon.github.io/EnvEcoStat/exe/exe_B.html).
:::

## The epidemiological problem

`DDT` is extremely effective against malaria mosquitoes and is therefore widely
used in areas where malaria is endemic.

At the same time, `DDT` could represent a [health risk]{.orange}, especially in the
case of [pregnant women]{.blue}.

For a sample of $n = 2312$ pregnant women, `DDE` is measured, that is a
substance related to `DDT`, present in the maternal serum during the third
trimester of pregnancy.

The variable `GAD` (Gestational Age at Delivery) measures instead on which day
of the pregnancy the delivery took place.

::: callout-note
#### Research question

Is the quantity of `DDE` higher among women who delivered prematurely?
:::

## Importing the `dde` data

To be able to run the next commands, you need to download the file `dde.csv` and
save it on your computer. [Link to the file](data/dde.csv)

```{r}
dde <- read.table("data/dde.csv", header = TRUE, sep = ",")
```

Alternatively, we can simply obtain them using the link:

```{r}
#| eval: false
dde <- read.table("https://tommasorigon.github.io/EnvEcoStat/data/dde.csv",
  header = TRUE, sep = ","
) # Downloads the file from the internet
```

```{r}
str(dde)
```

Note the presence of a new [type of variable]{.orange}, that is `integer`, namely
[integer numbers]{.blue}.

## Preliminary operations

A total of $361$ women [delivered prematurely]{.orange}, that is before the end of the
37th week.

To check this, we split the variable `DDE` into two groups.

```{r}
dde_preterm <- dde$DDE[dde$GAD < 37 * 7] # Preterm delivery group
dde_non_preterm <- dde$DDE[dde$GAD >= 37 * 7] # Non-preterm delivery group

length(dde_preterm) # Sample size of group 1
length(dde_non_preterm) # Sample size of group 2
```

The first group (variable `dde_preterm`) refers to preterm deliveries and
comprises $361$ observations.

The second group (variable `dde_non_preterm`) considers non-preterm deliveries
and comprises the remaining $1951$ observations.

## Splitting into intervals

To summarise the data through [frequencies]{.blue}, we can split the interval that
contains all the observed values (that is $(0,180]$) into a certain number of
[sub-intervals]{.orange}.

First of all, we define the $10$ sub-intervals of length $18$, closed on the
right, through the command `cut`:

```{r}
breaks <- 18 * (0:10) # Definition of the intervals: 10 intervals of length 18

dde_preterm_class <- cut(dde_preterm, breaks = breaks)
dde_non_preterm_class <- cut(dde_non_preterm, breaks = breaks)

head(dde_preterm_class)
```

In this way, we have transformed a `numeric` variable into a `factor` variable.

## Absolute frequencies

The [absolute frequencies]{.blue} $n_1,\dots, n_k$ are obtained through the command
`table`, which makes sense to apply to variables of type `factor`.

The function `table` [counts]{.orange} how many times a given category appears in the
vector.

We perform this operation for both groups of data:

```{r}
freq_abs_preterm <- table(dde_preterm_class)
freq_abs_preterm

freq_abs_non_preterm <- table(dde_non_preterm_class)
freq_abs_non_preterm
```

## Relative frequencies

Once the absolute frequencies have been obtained, it is possible to compute the
[relative frequencies]{.blue}.

Although it is possible to use the command `prop.table` (we will meet it again
later), here we follow a more direct route.

Recall indeed that $n = n_1 + \cdots + n_k$, therefore we can obtain the
relative frequencies $f_j = n_j / n$ as follows:

```{r}
freq_rel_preterm <- freq_abs_preterm / sum(freq_abs_preterm)
round(freq_rel_preterm, digits = 3)

freq_rel_non_preterm <- freq_abs_non_preterm / sum(freq_abs_non_preterm)
round(freq_rel_non_preterm, digits = 3)
```

## Organising the results

We then collect the results just obtained inside a matrix, to display them
better.

```{r}
tab_summary <- cbind(
  freq_abs_preterm, freq_abs_non_preterm,
  freq_rel_preterm, freq_rel_non_preterm
)

colnames(tab_summary) <- c(
  "n_j preterm",
  "n_j non-preterm",
  "f_j preterm",
  "f_j non-preterm"
)

round(tab_summary, 3) # Display of the results
```

## Histograms I

The [histogram]{.orange} makes it possible to represent a frequency distribution
graphically. To refresh your memory, recall that in its simplest version one
sets:

$$
\begin{aligned}
(\text{base of the rectangles}) &= (\text{length of the intervals}) \\
(\text{height of the rectangles}) &= (\text{absolute frequencies}) 
\end{aligned}
$$

This definition is [not appropriate]{.blue} if the [intervals]{.orange} have **different
sizes**.

In that case, the heights of the rectangles must be [proportional]{.blue} to the
[density]{.orange} of the observations in the individual classes.

To sum up, we will build the histograms setting

$$
\begin{aligned}
(\text{base of the rectangles}) &= (\text{length of the intervals}) \\
(\text{height of the rectangles}) &= \lambda \times (\text{density}) = \lambda \times \frac{(\text{absolute frequencies}) }{(\text{length of the intervals})} 
\end{aligned}
$$

where one typically sets $\lambda = 1/n$.

## Histograms II

Let us initially focus only on the group of [non-preterm births]{.blue}, to
understand how the **R** syntax works.

```{r}
par(mfrow = c(1, 2)) # Splits the plot into two parts

# First plot, absolute frequencies
hist(dde_non_preterm,
  freq = TRUE,
  breaks = 10, # We use 10 sub-intervals
  # From here on we are only adding aesthetic details
  main = "Non-preterm births",
  xlab = "DDE",
  ylab = "Absolute frequencies"
)

# Second plot, density
hist(dde_non_preterm,
  freq = FALSE, # Frequencies are NOT used
  breaks = 10, # We use 10 sub-intervals
  # From here on we are only adding aesthetic details
  main = "Non-preterm births",
  xlab = "DDE",
  ylab = "Density"
)
```

Histogram based on the [absolute frequencies]{.orange} (left) and on the [density]{.blue}
(right), only for the group of non-preterm births.

## Histograms III

::: callout-warning
In the case of [unequally spaced]{.orange} intervals, it is mandatory to use the option
`freq = FALSE`, otherwise the resulting plot is meaningless.
:::

```{r}
# Definition of UNEQUALLY spaced intervals
breaks <- c(5 * 0:10, 70, 90, 110, 130, 150, 170, 190)

# CORRECT PLOT
hist(dde_non_preterm,
  freq = FALSE,
  breaks = breaks,
  # From here on we are only adding aesthetic details
  main = "CORRECT PLOT", xlab = "DDE", ylab = "Density"
)

# WRONG PLOT
hist(dde_non_preterm,
  freq = TRUE, # Frequencies are used
  breaks = breaks,
  # From here on we are only adding aesthetic details
  main = "WRONG PLOT", xlab = "DDE", ylab = "Absolute frequencies"
)
```

The plot on the left is correct and makes use of the notion of [density]{.blue}
explained above. The plot on the right is [wrong]{.orange}.

Fortunately **R** warns us through a `warning` that this procedure is incorrect.

## The number of intervals

A number of intervals that is [too low]{.blue} entails a [loss of information]{.orange}.

Conversely, a number of intervals that is [too high]{.blue} entails a **loss of
synthesis**.

Keep in mind that the [number of intervals]{.orange} must depend on the **number of
data points**.

Various formulas have been proposed to identify the "optimal number" of
intervals. They should however be taken as suggestions and not used
automatically.

[Sturges]{.blue}. The number of intervals, rounded to the nearest integer, is $$ 
(\text{number of intervals}) = 1 + \log_2{n}. 
$$

[Freedman & Diaconis]{.orange}. The number of intervals, rounded to the nearest
integer, is $$ 
(\text{number of intervals})  = \frac{x_{(n)} - x_{(1)}}{2(\mathcal{Q}_{0.75} - \mathcal{Q}_{0.25})}n^{1/3}. 
$$

## The number of intervals II

The "optimal" number of intervals can be obtained through the following
commands:

```{r}
nclass.Sturges(dde_non_preterm) # Sturges' rule
nclass.FD(dde_non_preterm) # Freedman and Diaconis' rule
```

::: callout-note
#### Note

The **R** command `hist`, unless specified otherwise through the option
`breaks`, automatically selects [equally spaced]{.blue} intervals through the rule of
[Sturges]{.orange}.
:::

## Comparison between the two distributions

This final plot finally makes it possible to compare the two distributions.

```{r}
hist(dde_non_preterm, # Non-preterm births group
  freq = FALSE,
  main = "Non-preterm births"
)

hist(dde_preterm, # Preterm births group
  freq = FALSE,
  main = "Preterm births"
)
```

Possibly, it is possible to select a different rule by setting `breaks = "FD"`.

The distribution of "preterm births" turns out to be shifted further to the
right.

## The empirical cumulative distribution function I

A second graphical representation of frequent use is the so-called **empirical
cumulative distribution function** $F(x)$.

Let $x_1,\dots,x_n$ be a collection of data, then we define $$
F(x) = \frac{1}{n} \sum_{i=1}^n \mathbb{1}(x_i \le x),
$$ where $\mathbb{1}(x_i \le x)$ is the [indicator function]{.blue} and equals $1$ if
$x_i \le x$ and $0$ if $x_i > x$.

In **R**, one uses the command `ecdf` (*empirical cumulative distribution
function*), for example:

```{r}
ecdf(dde_non_preterm)
```

## The empirical cumulative distribution function II

The command `ecdf` returns as an object an actual [function]{.orange}. Indeed:

```{r}
F_non_preterm <- ecdf(dde_non_preterm)
F_non_preterm(c(20, 40, 60)) # Evaluates the empirical cdf at 20, 40 and 60
sum(dde_non_preterm <= 40) / length(dde_non_preterm) # Alternative "manual" command
```

To draw a plot of it, it is moreover enough to use the command `plot`:

```{r}
par(mfrow = c(1, 1))
plot(ecdf(dde_non_preterm))
```

## The empirical cumulative distribution function III

To be able to compare the cumulative distribution functions of the two groups,
the syntax is slightly more elaborate:

```{r}
plot(ecdf(dde_preterm),
  do.points = FALSE, col = "blue",
  main = "Blue: preterm delivery. Red: non-preterm delivery",
  xlab = "DDE",
  ylab = "F(DDE)"
)

plot(ecdf(dde_non_preterm), col = "red", add = TRUE) # Add a second group
```

The option `do.points = FALSE` omits the "dots" in the plot and has been added
for purely [aesthetic]{.blue} reasons.

## Location indices: the mean

To quantify the difference between the two distributions, we want to find the
[arithmetic means]{.orange} of the variables `dde_preterm` and `dde_non_preterm`.

In **R** there is of course a dedicated command:

```{r}
mean(dde_preterm) # Arithmetic mean of DDE for women with preterm delivery
mean(dde_non_preterm) # Arithmetic mean of DDE for women with regular delivery
```

As already widely noted, we observe that DDE is present in larger quantities
among the women who delivered prematurely.

The arithmetic mean could also have been obtained starting from the commands we
met in the previous units:

```{r}
n <- length(dde_preterm) # Sample size of dde_preterm (= 361)
sum(dde_preterm) / n # Arithmetic mean of the variable dde_preterm
```

## Location indices: the median I

The median (Me) of a variable is the [central value]{.blue} of the distribution, that
is:

$$
\text{Me} = \begin{cases}x_{\left(\frac{n+1}{2}\right)}, &\qquad \text{ if } n \text{ is odd},  \\[10pt] \left(x_{(n/2)} + x_{(n/2+1)}\right)/2,&\qquad \text{ if } n \text{ is even},\end{cases}
$$ where $x_{(1)},\dots,x_{(n)}$ represents the ordered sample.

In **R** there is a dedicated command:

```{r}
median(dde_preterm) # Median of DDE for women with preterm delivery
median(dde_non_preterm) # Median of DDE for women with regular delivery
```

Note (why?) that the empirical cumulative distribution function, evaluated at
the median, is approximately equal to `0.5`.

```{r}
F_non_preterm(median(dde_non_preterm))
```

## Location indices: the median II

The median of a numerical variable can also be obtained "manually".

```{r}
x <- c(10, 20, 25, 3.5, 28, 62)
n <- length(x) # Sample size
n # Note that n = 6 is even

x_sort <- sort(x) # Ordered vector of the values of x
x_sort

pos_med_1 <- n / 2 # Element in position n/2
pos_med_2 <- n / 2 + 1 # Element in position n/2+1

(x_sort[pos_med_1] + x_sort[pos_med_2]) / 2 # Median of x

median(x) # Of course, the result must coincide with:
```

::: callout-note
#### Exercise

Obtain the median "manually" in the case in which the vector `x` has an odd
number of elements.
:::

## Location indices: the quantiles I

Let $x_1,\dots,x_n$ be a set of data, let $p \in (0,1)$ and let $F(x)$ be the
empirical cumulative distribution function. The [$p$-quantile]{.orange} is then equal
to

$$
\mathcal{Q}_p = \inf\{ x : F(x) \ge p \}.
$$

In **R**, considering $p = (0.1,0.25,0.75,0.9)$, that is the first decile, the
first quartile, the third quartile and the ninth decile, we can use the command
`quantile`

```{r}
quantile(dde_preterm, probs = c(0.1, 0.25, 0.75, 0.9), type = 1)
quantile(dde_non_preterm, probs = c(0.1, 0.25, 0.75, 0.9), type = 1)
```

::: callout-note
#### Exercise

Convince yourself that the "manual" implementation of the command `quantile` is
the following:

```{r}
min(dde_non_preterm[F_non_preterm(dde_non_preterm) >= 0.25])
min(dde_non_preterm[F_non_preterm(dde_non_preterm) >= 0.75])
```
:::

## Location indices: the quantiles II

::: callout-note
#### Note

The definition of quantile we have given is the usual one, but it is not the
only possible one.
:::

That definition is obtained with the option `type = 1`, which however is **not
the default**. The default value is instead `type = 7`.

If you are curious to know the various definitions of quantile, remember that
the [documentation]{.blue} can be consulted through the command: `? quantile`.

::: callout-note
#### Note

A drawback of the `type = 1` quantile is that the median does not always
coincide with the quantile $\mathcal{Q}_{0.5}$. Find an example in which this
happens and check that the problem is instead solved in the `type = 7` case.
:::

## Location indices: the quantiles III

In **R** there are in total as many as [9 ways]{.orange} of computing the quantiles.

Fortunately, for large sample sizes the [differences]{.blue} tend to be
[negligible]{.orange}.

We list below some examples for the variable `dde_preterm`:

```{r}
tab <- rbind(
  quantile(dde_preterm, probs = c(0.1, 0.25, 0.5, 0.75, 0.9), type = 1),
  quantile(dde_preterm, probs = c(0.1, 0.25, 0.5, 0.75, 0.9), type = 6),
  quantile(dde_preterm, probs = c(0.1, 0.25, 0.5, 0.75, 0.9), type = 7),
  quantile(dde_preterm, probs = c(0.1, 0.25, 0.5, 0.75, 0.9), type = 9)
)
rownames(tab) <- c(1, 6, 7, 9) # Changes the names of the rows of the table
tab
```

::: callout-tip
#### Hint

What does `rbind` do? We met it in Unit A.
:::

## The `summary` command

To obtain quickly the [main descriptive statistics]{.blue} of a distribution
(minimum, maximum, mean, median and quartiles), one often uses the command
`summary`:

```{r}
summary(dde_preterm)
summary(dde_non_preterm)
```

The command `summary` can also be used directly on an object of type
`data.frame`, producing the following output:

```{r}
summary(dde)
```

## Boxplots

An alternative graphical representation to histograms are [boxplots]{.orange}.

The [descriptive statistics]{.blue} on which boxplots are based can be obtained
through the command `boxplot.stats`, that is:

```{r}
boxplot.stats(dde_preterm)
```

```{r}
boxplot(dde_preterm, dde_non_preterm) # Produces the plot
```
