---
title: "Descriptive analysis of the hemoglobin data"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
page-layout: full
bibliography: biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    smooth-scroll: true
    fig-dpi: 90
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

## [Homepage](index.html)

```{r}
#| warning: false
#| echo: false
#| include: false
#| message: false
#| purl: false

# Please ignore this chunk.
# A single purl() call: calling it twice in the same knit would clash on the
# chunk labels. The clean script is derived from the full one, so the two
# cannot go out of sync.

# 1. Full script, for personal use: code plus the text of the notes as
#    comments. Not linked from the website (and ignored by git).
knitr::purl("un_E.qmd", output = "code/un_E_full.R", documentation = 2)
full <- readLines("code/un_E_full.R")
# strip the Quarto markup that is only meaningful in the rendered page
full <- full[!grepl("^#' *:::", full)]
full <- gsub("\\[([^]]*)\\]\\{[^}]*\\}", "\\1", full)
squeeze <- function(x) x[!(x == "" & c(FALSE, head(x == "", -1)))]
writeLines(squeeze(full), "code/un_E_full.R")

# 2. Clean script, linked from the website: code only, obtained by dropping
#    the text (#') and the chunk headers (#+).
clean <- squeeze(full[!grepl("^#'|^#\\+", full)])
writeLines(clean, "code/un_E.R")
styler:::style_file("code/un_E.R")
```

## Topics covered

-   Variance and standard deviation
-   Other measures of variability (range, interquartile range, MAD)

## Description of the problem

We are interested in comparing the effectiveness of two different methods,
called A and B, for the [measurement of haemoglobin]{.orange} in blood.

[Artificial blood]{.blue} containing 15 grams of haemoglobin every
$100 \text{cm}^3$ has been created in the laboratory.

A total of $n = 360$ samples were drawn from the compound.

Of these, in $n_A = 180$ samples the haemoglobin was measured using method A,
whereas for the remaining $n_B = 180$ samples method B was used.

The differences between the various measurements are largely to be attributed to
the measurement errors of the two different methods.

## Importing the `hemoglobin` data

As done previously, first of all you need to download the file `hemoglobin.csv`
and save it on your computer. [Link to the file](data/hemoglobin.csv)

```{r}
hemoglobin <- read.table("data/hemoglobin.csv",
  header = TRUE, sep = ",", stringsAsFactors = TRUE
)
```

Alternatively, we can simply obtain them using the link:

```{r}
#| eval: false
path <- "https://tommasorigon.github.io/EnvEcoStat/data/hemoglobin.csv"
hemoglobin <- read.table(path, header = TRUE, sep = ",", stringsAsFactors = TRUE)
```

```{r}
str(hemoglobin)
```

The option `stringsAsFactors = TRUE` implies that the variable `method` is coded
as a `factor` and not as a `character`.

## Preliminary operations

In this format the data are [hard to analyse]{.orange}, at least without using the
advanced `*apply` functions or the tools of the so-called `tidyverse`.

Therefore, we create the two variables `hemo_A` and `hemo_B`, containing the
haemoglobin values for the two methods.

```{r}
hemo_A <- hemoglobin$hemoglobin[hemoglobin$method == "A"] # Group A
hemo_B <- hemoglobin$hemoglobin[hemoglobin$method == "B"] # Group B

summary(hemo_A)
summary(hemo_B)
```

```{r}
boxplot(hemoglobin$hemoglobin ~ hemoglobin$method,
  ylab = "Haemoglobin",
  xlab = "Method"
)
```

From these first descriptive analyses it emerges that both methods are, at least
on average, [well calibrated]{.blue}. Which of the two is preferable?

## The variance I

Recall that the [variance]{.orange} of the data $x_1,\dots,x_n$ is equal to

$$
\sigma^2  = \frac{1}{n}\sum_{i=1}^n (x_i - \bar{x})^2 = \left(\frac{1}{n}\sum_{i=1}^n x_i^2\right) - \bar{x}^2.
$$

We can therefore compute the variances of the two groups "manually":

```{r}
mean((hemo_A - mean(hemo_A))^2) # Variance of group A
mean((hemo_B - mean(hemo_B))^2) # Variance of group B

mean(hemo_A^2) - mean(hemo_A)^2 # Variance of group A, alternative formula
mean(hemo_B^2) - mean(hemo_B)^2 # Variance of group B, alternative formula
```

Method B is therefore characterised by a larger [variability]{.blue}.

## The variance II

For convenience, we create a [new function]{.orange} called `my_var` that computes the
[variance]{.blue} of a generic vector `x`. We will therefore have:

```{r}
my_var <- function(x) {
  mean(x^2) - mean(x)^2
}

my_var(hemo_A)
my_var(hemo_B)
```

In **R** the function `var` is available, however this computes a slightly
[different quantity]{.orange}, that is $$
\texttt{var(x)} = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2,
$$ that is the [unbiased estimator]{.blue} of the variance.

[Not clear why?]{.orange} It will be addressed in the part of the course devoted to
statistical inference.

## The variance III

The difference between `my_var` and `var` is, in practice, essentially
[negligible]{.blue} in this case, indeed:

```{r}
var(hemo_A)
var(hemo_B)
```

::: callout-note
#### Exercise

Obtain the same output as `my_var` using the function `var`.
:::

## Standard deviation

The [standard deviation]{.orange} is the square root of the variance, that is $$
\text{sd}(x) = \sigma = \sqrt{\sigma^2}.
$$

We therefore create a suitable **R** function called `my_sd` to compute it.

Also in this case, be careful that the **R** command `sd` computes instead the
value of $$
\texttt{sd(x)} = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2}.
$$

```{r}
my_sd <- function(x) {
  sqrt(my_var(x))
}

my_sd(hemo_A)
my_sd(hemo_B)

sd(hemo_A)
sd(hemo_B)
```

## Range

The [range]{.blue} is the difference between the minimum and the maximum of the
distribution, that is $$
(\text{``Range''}) = x_{(n)} - x_{(1)},
$$ where $x_{(1)}$ and $x_{(n)}$ represent respectively the minimum and the
maximum of the data.

Although a specific function to compute it does not exist, we can use the
following commands

```{r}
max(hemo_A) - min(hemo_A)
max(hemo_B) - min(hemo_B)

diff(range(hemo_A))
diff(range(hemo_B))
```

::: callout-note
#### Exercise

Define a suitable **R** function that computes the [range]{.orange}.
:::

## Interquartile range

The [interquartile range]{.blue} is the difference between the third and the first
quartile, that is $$
(\text{``Interquartile range''}) = \mathcal{Q}_{0.75} - \mathcal{Q}_{0.25}.
$$

It is much more [resistant]{.orange} than the variance in the presence of a few extreme
observations.

Its implementation in **R** is the following:

```{r}
interquartile_range <- function(x) {
  diff(quantile(x, probs = c(0.25, 0.75)))
}

interquartile_range(hemo_A)
interquartile_range(hemo_B)
```

## Median absolute deviation (MAD)

The variability index [MAD]{.blue} (*Median Absolute Deviation*) is defined as
follows $$
\text{MAD} = \text{Median}(|x_1 - \text{Me}_x|, \dots, |x_n - \text{Me}_x|), \qquad \text{Me}_x = \text{Median}(x_1,\dots,x_n).
$$

To compute it, we create an appropriate function:

```{r}
MAD <- function(x) {
  median(abs(x - median(x)))
}
```

The MAD of the variables `hemo_A` and `hemo_B` is therefore equal to

```{r}
MAD(hemo_A)
MAD(hemo_B)
```

## The `tapply` function I

There exists a more [elegant]{.orange} and [fast]{.blue} method to obtain these results,
based on the function `tapply`.

The function `tapply(x, group, fun)` applies a certain function `fun` to a set
of data `x`, for each [group]{.orange} `group`.

For example, the arithmetic mean of `hemo_A` and `hemo_B` is obtained as
follows:

```{r}
tapply(hemoglobin$hemoglobin, hemoglobin$method, mean)
with(hemoglobin, tapply(hemoglobin, method, mean)) # Even more compactly
```

## The `tapply` function II

Therefore, we can obtain the various variability indices as follows:

```{r}
tab <- rbind(
  with(hemoglobin, tapply(hemoglobin, method, my_var)),
  with(hemoglobin, tapply(hemoglobin, method, my_sd)),
  with(hemoglobin, tapply(hemoglobin, method, interquartile_range)),
  with(hemoglobin, tapply(hemoglobin, method, MAD))
)
rownames(tab) <- c("Variance", "Standard deviation", "Interquartile range", "MAD")
tab
```

::: callout-note
#### Exercise

Try to use the function `aggregate` to obtain the same result.
:::

## Summary exercise

The most commonly used measure of skewness is the so-called standardised
[skewness index]{.blue} of [Pearson]{.orange}, defined as $$
\gamma = \frac{1}{\text{sd}(x)^3} \frac{1}{n} \sum_{i=1}^n (x_i - \bar{x})^3 = \frac{1}{n} \sum_{i=1}^n \left(\frac{x_i - \bar{x}}{\sigma}\right)^3.
$$

-   Create a suitable function to compute it.

#### Solution of the summary exercise

```{r}
# First possible implementation
asym <- function(x) {
  sdev <- sqrt(mean(x^2) - mean(x)^2)
  mean((x - mean(x))^3) / sdev^3
}

# (Slightly) alternative implementation
asym <- function(x) {
  sdev <- sqrt(mean(x^2) - mean(x)^2)
  z <- (x - mean(x)) / sdev
  mean(z^3)
}

asym(hemo_A)
```
