---
title: "Descriptive analysis of the forbes data"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
page-layout: full
bibliography: biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    smooth-scroll: true
    fig-dpi: 90
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

## [Homepage](index.html)

```{r}
#| warning: false
#| echo: false
#| include: false
#| message: false
#| purl: false

# Please ignore this chunk.
# A single purl() call: calling it twice in the same knit would clash on the
# chunk labels. The clean script is derived from the full one, so the two
# cannot go out of sync.

# 1. Full script, for personal use: code plus the text of the notes as
#    comments. Not linked from the website (and ignored by git).
knitr::purl("un_F.qmd", output = "code/un_F_full.R", documentation = 2)
full <- readLines("code/un_F_full.R")
# strip the Quarto markup that is only meaningful in the rendered page
full <- full[!grepl("^#' *:::", full)]
full <- gsub("\\[([^]]*)\\]\\{[^}]*\\}", "\\1", full)
squeeze <- function(x) x[!(x == "" & c(FALSE, head(x == "", -1)))]
writeLines(squeeze(full), "code/un_F_full.R")

# 2. Clean script, linked from the website: code only, obtained by dropping
#    the text (#') and the chunk headers (#+).
clean <- squeeze(full[!grepl("^#'|^#\\+", full)])
writeLines(clean, "code/un_F.R")
styler:::style_file("code/un_F.R")
```

## Topics covered

-   Covariance and correlation
-   The simple linear regression model
-   Least squares and goodness of fit

## Description of the problem

For $n = 17$ locations in the Alps, the [atmospheric pressure]{.orange} (inHg, that is
"*inches of mercury*") and the [boiling temperature]{.blue} of water (in degrees
Fahrenheit) are measured.

The data come from an experiment carried out by the **Scottish physicist
Forbes** in 1857.

Forbes was interested in estimating the altitude through the pressure. However,
at the time the barometer was a heavy and expensive instrument.

In the mountains water boils at a different temperature, so that one can try to
estimate the pressure starting from the boiling temperature.

## Importing the `forbes` data

As done previously, first of all you need to download the file `forbes.csv` and
save it on your computer. [Link to the file](data/forbes.csv)

```{r}
forbes <- read.table("data/forbes.csv", header = TRUE, sep = ",")
```

Alternatively, we can simply obtain them using the link:

```{r}
#| eval: false
path <- "https://tommasorigon.github.io/EnvEcoStat/data/forbes.csv"
forbes <- read.table(path, header = TRUE, sep = ",")
```

```{r}
str(forbes)
```

## Preliminary operations

For [interpretative]{.orange} reasons, we convert the temperature from degrees
Fahrenheit to degrees Celsius, recalling that

$$
(\text{``Fahrenheit''}) = 32 + \frac{9}{5}(\text{``Celsius''}).
$$

```{r}
colnames(forbes) <- c("TempF", "Pressure") # Change the names of the variables

forbes$TempC <- round((forbes$TempF - 32) * 5 / 9, 2) # From Fahrenheit to Celsius
summary(forbes)
```

Why do we round the values using `round`? For [aesthetic]{.blue} reasons only.

## Histogram

::: callout-warning
#### Question 1

Draw a histogram of the temperature variable, choosing an appropriate number of
equally spaced classes and justifying the choice.

Add to these plots the graphical "embellishments" deemed necessary (names of the
variables, title, and so on).
:::

#### Solution

We can decide to specify the intervals of the classes ourselves, or to leave
this choice to **R**.

```{r}
par(mfrow = c(1, 2)) # Split the graphical window into 2 parts

# Option 1, for a total of 6 equally spaced classes
hist(forbes$TempC) # Equivalent to: hist(forbes$TempC, breaks = "sturges")

# Option 2, we define manually 5 equally spaced classes
breaks <- c(90, 92.5, 95, 97.5, 100, 102.5)
hist(forbes$TempC, breaks = breaks)
```

The solution on the left uses $6$ classes. The one on the right uses $5$
classes.

## Location indices

::: callout-warning
#### Question 2

Obtain the arithmetic mean of both variables `TempC` and `Pressure`. What is the
average boiling temperature expressed in degrees Fahrenheit? Answer without
computing all the values of the variable `TempF`.
:::

#### Solution

We have already computed the means through the command `summary`; for
completeness:

```{r}
# First part of the question
mean(forbes$TempC)
mean(forbes$Pressure)
```

```{r}
# Second part of the question
32 + 9 / 5 * mean(forbes$TempC) # Using the properties of the mean
mean(forbes$TempF) # Not required: computes the mean from the original data
```

::: callout-note
#### Exercise

Why do the means computed in the second part differ slightly? What can this be
due to?
:::

## Variability indices

::: callout-warning
#### Question 3

Obtain the variance of the variables `TempC` and `Pressure`.
:::

#### Solution

Since it will turn out to be useful later, we define the function `my_var` that
computes the [variance]{.orange}.

```{r}
# Yes, the function is defined on a single line and there is nothing wrong with it
my_var <- function(x) mean(x^2) - mean(x)^2

# Computation of the two variances
my_var(forbes$TempC)
my_var(forbes$Pressure)
```

::: callout-note
#### Exercise

Obtain the second moments of the variables `TempC` and `Pressure`.
:::

## Covariance and correlation

::: callout-warning
## Question 4

Draw a suitable plot that helps to understand the relationship between the two
variables. Then compute the correlation.
:::

#### Solution

For the first part of the question, we need the new command `plot`, which can be
used (among other things!) to build a [scatterplot]{.blue}.

We provide two versions of the same plot; the second one contains aesthetic
improvements.

```{r}
par(mfrow = c(1, 1)) # We want to show one plot at a time
plot(forbes$TempC, forbes$Pressure)
plot(forbes$TempC, forbes$Pressure,
  pch = 16,
  xlab = "Temperature", ylab = "Pressure"
)
```

It is therefore evident that the data are approximately (although not perfectly)
aligned.

For the second part of the question (correlation), we first have to obtain the
[covariance]{.orange} between two variables.

The [covariance]{.blue} between two sets of data $x_1,\dots,x_n$ and $y_1,\dots,y_n$
is defined as

$$
\text{cov}(x,y) = \frac{1}{n}\sum_{i=1}^n(x_i - \bar{x})(y_i - \bar{y}) = \frac{1}{n}\sum_{i=1}^nx_i y_i - \bar{x}\bar{y}.
$$

We therefore define the function `my_cov`, which computes precisely the
covariance:

```{r}
my_cov <- function(x, y) mean(x * y) - mean(x) * mean(y)
my_cov(forbes$TempC, forbes$Pressure) # = my_cov(forbes$Pressure, forbes$TempC)
```

In **R** there is also the command `cov` which, as in the case of the variance,
divides the sum by $(n - 1)$ and not by $n$, for reasons related to
[statistical inference]{.orange}:

```{r}
cov(forbes$Pressure, forbes$TempC) # = 17 / 16 * my_cov(forbes$TempC, forbes$Pressure)
```

The [correlation]{.blue} index is defined as: $$
\rho = \frac{\text{cov}(x,y)}{\sqrt{\text{var}(x) \text{var}(y)}}.
$$

Therefore, we can compute the correlation as follows:

```{r}
my_cov(forbes$TempC, forbes$Pressure) /
  sqrt(my_var(forbes$TempC) * my_var(forbes$Pressure))

cov(forbes$TempC, forbes$Pressure) /
  sqrt(var(forbes$TempC) * var(forbes$Pressure))
```

::: callout-note
#### Exercise

Why do the results of the two commands coincide? Check this fact by carrying out
the computations [analytically]{.orange} (pen and paper).
:::

In **R** there is also the command `cor`, which returns the correlation

```{r}
correlation <- cor(forbes$TempC, forbes$Pressure)
correlation
```

## The linear regression model

::: callout-warning
## Question 5

Obtain the least squares line for the relationship between `TempC` and
`Pressure` and draw it in the plot obtained previously.
:::

#### Solution

First of all recall that in a [linear model]{.blue} of the form
$y_i = \alpha + \beta x_i + \epsilon_i$, the [least squares]{.orange} estimates are
equal to $$
\hat{\alpha} = \bar{y} - \hat{\beta}\:\bar{x}, \qquad \hat{\beta}  = \frac{\text{cov}(x,y)}{\text{var}(x)}.
$$

Therefore, we can compute them as follows:

```{r}
# Slope
beta_hat <- my_cov(forbes$TempC, forbes$Pressure) / my_var(forbes$TempC)
# Intercept
alpha_hat <- mean(forbes$Pressure) - mean(forbes$TempC) * beta_hat

c(alpha_hat, beta_hat)
```

```{r}
plot(forbes$TempC, forbes$Pressure,
  pch = 16,
  xlab = "Temperature", ylab = "Pressure"
)
abline(a = alpha_hat, b = beta_hat)
```

::: callout-note
#### Note

The same estimates are obtained with the command `lm`, which we will meet again
in the part of the course devoted to linear models:

```{r}
lm(Pressure ~ TempC, data = forbes)
```
:::

::: callout-warning
## Question 6

According to the estimated model, if the boiling temperature of water is equal
to 97 degrees Celsius, what is the pressure?
:::

#### Solution

Using the estimates obtained, we can quickly compute the fitted values:

```{r}
x <- seq(from = 90, to = 100, length = 20)
alpha_hat + beta_hat * x
```

In particular, when `TempC = 97` we have that:

```{r}
alpha_hat + beta_hat * 97
```

::: callout-warning
## Question 7

Obtain an index of goodness of fit of the curve to the data.
:::

#### Solution

The coefficient $R^2$ for a simple linear regression model is defined as: $$
R^2 = 1 - \frac{\text{var}(r)}{\text{var}(y)} = \rho^2,
$$ where $r_1,\dots,r_n$ are the [residuals]{.blue}.

First of all we therefore compute the residuals:

```{r}
residuals <- forbes$Pressure - (alpha_hat + beta_hat * forbes$TempC)
```

The coefficient $R^2$ can then be obtained in two different ways:

```{r}
correlation^2
1 - my_var(residuals) / my_var(forbes$Pressure)
```

## Summary exercise

The following [synthetic]{.orange} data report, for $10$ hypothetical countries,
the per capita income (thousands of €, `income`) and the per capita electricity
consumption (MWh, `consumption`) in a given year. They are made up for the
purpose of this exercise and are not real statistics.

```{r}
income <- c(18.2, 21.4, 25.0, 27.6, 30.1, 33.4, 36.8, 41.2, 45.9, 52.3)
consumption <- c(3.1, 3.6, 4.0, 4.2, 4.9, 5.1, 5.8, 6.0, 6.9, 7.4)
```

-   Draw a scatterplot of the two variables and comment on it.

-   Compute the covariance and the correlation between `income` and
    `consumption`.

-   Obtain the least squares line that explains the consumption as a function of
    the income, and add it to the plot.

-   Compute the $R^2$ and comment on the goodness of fit.

-   According to the estimated model, what is the expected consumption for a
    country with a per capita income of $40$ thousand €?

::: callout-tip
#### Hint

To import small amounts of data into **R**, it is also possible to use the
command `scan`.
:::
