---
title: "Descriptive analysis of the titanic data"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
page-layout: full
bibliography: biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    smooth-scroll: true
    fig-dpi: 90
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

## [Homepage](index.html)

```{r}
#| warning: false
#| echo: false
#| include: false
#| message: false
#| purl: false

# Please ignore this chunk.
# A single purl() call: calling it twice in the same knit would clash on the
# chunk labels. The clean script is derived from the full one, so the two
# cannot go out of sync.

# 1. Full script, for personal use: code plus the text of the notes as
#    comments. Not linked from the website (and ignored by git).
knitr::purl("un_G.qmd", output = "code/un_G_full.R", documentation = 2)
full <- readLines("code/un_G_full.R")
# strip the Quarto markup that is only meaningful in the rendered page
full <- full[!grepl("^#' *:::", full)]
full <- gsub("\\[([^]]*)\\]\\{[^}]*\\}", "\\1", full)
squeeze <- function(x) x[!(x == "" & c(FALSE, head(x == "", -1)))]
writeLines(squeeze(full), "code/un_G_full.R")

# 2. Clean script, linked from the website: code only, obtained by dropping
#    the text (#') and the chunk headers (#+).
clean <- squeeze(full[!grepl("^#'|^#\\+", full)])
writeLines(clean, "code/un_G.R")
styler:::style_file("code/un_G.R")
```

## Topics covered

-   Qualitative data
-   Contingency tables
-   Conditional distributions
-   Independence and the $\chi^2$ index of association

## Description of the problem

After the [Titanic disaster]{.orange}, a commission of inquiry of the *British Board of
Trade* compiled a list of all the 1316 passengers, including the following
information:

-   the outcome (saved, not saved)
-   the class (I, II, III) in which they travelled
-   sex, age, and so on.

In this unit we limit ourselves to considering the information on the
`survived` outcome and the `class`.

## Importing the `titanic` data

As done previously, first of all you need to download the file `titanic.csv` and
save it on your computer. [Link to the file](data/titanic.csv)

```{r}
titanic <- read.table("data/titanic.csv",
  header = TRUE, sep = ",", stringsAsFactors = TRUE
)
```

Alternatively, we can simply obtain them using the link:

```{r}
#| eval: false
path <- "https://tommasorigon.github.io/EnvEcoStat/data/titanic.csv"
titanic <- read.table(path, header = TRUE, sep = ",", stringsAsFactors = TRUE)
```

```{r}
str(titanic)
```

## Absolute and relative (marginal) frequencies

We can obtain the (marginal) absolute frequencies of the two variables using the
command `summary`:

```{r}
summary(titanic)
```

Of course, we can obtain the [absolute]{.blue} and [relative]{.orange} frequencies also
using the command `table`. For example for the variable `class`, we can use:

```{r}
freq_abs_class <- table(titanic$class)
freq_rel_class <- freq_abs_class / sum(freq_abs_class)
tab_summary <- cbind(freq_abs_class, freq_rel_class)
tab_summary
```

## Joint frequencies

A summary we can produce consists in building a table, called a **contingency
table[ or ]{.blue}two-way table**.

In **R** one uses, also in this case, the command `table`, with two arguments:

```{r}
tab <- table(titanic$survived, titanic$class)
tab
```

In this table the [joint frequencies]{.orange} are reported: for example, the value
$203$ represents the number of passengers who travelled in first class and who
survived.

## Contingency table

Let $x$ and $y$ be two variables having categories $c_1,\dots,c_h$ and
$d_1,\dots,d_k$, respectively.

A [contingency table]{.blue} (for two variables) for the pairs of data
$(x_1,y_1),\dots,(x_n,y_n)$ takes the following form:

|                  |          |         | Variable $y$ |         |          |           |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| [Variable]{.orange} $x$ | $d_1$    | $\dots$ | $d_j$        | $\dots$ | $d_k$    | [Total]{.blue} |
| $c_1$            | $n_{11}$ | $\dots$ | $n_{1j}$     | $\dots$ | $n_{1k}$ | $n_{1+}$  |
| $\vdots$         | $\vdots$ |         | $\vdots$     |         | $\vdots$ | $\vdots$  |
| $c_i$            | $n_{i1}$ | $\dots$ | $n_{ij}$     | $\dots$ | $n_{ik}$ | $n_{i+}$  |
| $\vdots$         | $\vdots$ |         | $\vdots$     |         | $\vdots$ | $\vdots$  |
| $c_h$            | $n_{h1}$ | $\dots$ | $n_{hj}$     | $\dots$ | $n_{hk}$ | $n_{h+}$  |
| [Total]{.orange}        | $n_{+1}$ | $\dots$ | $n_{+j}$     | $\dots$ | $n_{+k}$ | $n$       |

The frequency $n_{ij}$ is the number of statistical units that simultaneously
present the categories $c_i$ and $d_j$.

## Contingency table, relative frequencies

Dividing each term of the previous table by $n$, one moreover obtains:

|                  |          |         | Variable $y$ |         |          |           |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| [Variable]{.blue} $x$ | $d_1$    | $\dots$ | $d_j$        | $\dots$ | $d_k$    | [Total]{.orange} |
| $c_1$            | $f_{11}$ | $\dots$ | $f_{1j}$     | $\dots$ | $f_{1k}$ | $f_{1+}$  |
| $\vdots$         | $\vdots$ |         | $\vdots$     |         | $\vdots$ | $\vdots$  |
| $c_i$            | $f_{i1}$ | $\dots$ | $f_{ij}$     | $\dots$ | $f_{ik}$ | $f_{i+}$  |
| $\vdots$         | $\vdots$ |         | $\vdots$     |         | $\vdots$ | $\vdots$  |
| $c_h$            | $f_{h1}$ | $\dots$ | $f_{hj}$     | $\dots$ | $f_{hk}$ | $f_{h+}$  |
| [Total]{.blue}        | $f_{+1}$ | $\dots$ | $f_{+j}$     | $\dots$ | $f_{+k}$ | $1$       |

The relative frequency $f_{ij} = n_{ij} / n$ is therefore the fraction of
observations that simultaneously present the categories $c_i$ and $d_j$.

## Joint & marginal frequencies

The tables described in the previous sections are obtained in **R** as follows:

```{r}
addmargins(tab) # Adds the (absolute) marginal distributions
```

```{r}
tab_rel <- prop.table(tab) # Alternative command: table(tab) / sum(tab)
tab_rel
```

```{r}
addmargins(tab_rel) # Adds the relative marginal distributions
```

## Conditional distributions I

#### Conditional distribution $(x \mid y = d_j)$

The $j$-th column shows the distribution of $x$ [conditional]{.orange} on $y = d_j$
or, equivalently, the distribution of $x$ given $y = d_j$.

| Distribution $x \mid y = d_j$ | $c_1$             | $\dots$ | $c_i$             | $\dots$ | $c_h$             | Total    |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Absolute frequencies          | $n_{1j}$          | $\dots$ | $n_{ij}$          | $\dots$ | $n_{hj}$          | $n_{+j}$ |
| Relative frequencies          | $n_{1j} / n_{+j}$ | $\dots$ | $n_{ij} / n_{+j}$ | $\dots$ | $n_{hj} / n_{+j}$ | $1$      |

#### Conditional distribution $(y \mid x = c_i)$

The $i$-th row shows the distribution of $y$ [conditional]{.blue} on $x = c_i$ or,
equivalently, the distribution of $y$ given $x = c_i$.

| Distribution $y \mid x = c_i$ | $d_1$             | $\dots$ | $d_j$             | $\dots$ | $d_k$             | Total    |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Absolute frequencies          | $n_{i1}$          | $\dots$ | $n_{ij}$          | $\dots$ | $n_{ik}$          | $n_{i+}$ |
| Relative frequencies          | $n_{i1} / n_{i+}$ | $\dots$ | $n_{ij} / n_{i+}$ | $\dots$ | $n_{ik} / n_{i+}$ | $1$      |

## Conditional distributions II

The command `prop.table` also makes it possible to compute the relative
conditional frequencies.

The distribution of each class, conditional on the outcome, is:

```{r}
prop.table(tab, 1)
```

The distribution of each outcome, conditional on the class, is:

```{r}
prop.table(tab, 2)
```

## Graphical representations

It is possible to represent the data of a contingency table graphically through
a bar chart, which in **R** is implemented through the command `barplot`.

A graphical representation of the joint frequencies is therefore the following:

```{r}
barplot(tab, beside = TRUE, legend.text = TRUE) # beside = TRUE places the bars side by side
```

If we were interested in showing the conditional frequencies, we can instead
use:

```{r}
barplot(prop.table(tab, 2),
  beside = FALSE,
  xlab = "Class",
  legend.text = TRUE
) # beside = FALSE stacks the bars.
```

## Advanced graphical tools (optional)

#### The `ggplot2` package

So-called **R** `base` contains a wide range of graphical representations.
Nevertheless, an **R** graphical package called `ggplot2` has recently gained
considerable popularity.

Let us first load it into memory:

```{r}
library(ggplot2)
```

Then, the first plot shown above can be obtained with the syntax:

```{r}
ggplot(data = titanic, aes(x = class, fill = survived)) +
  geom_bar(position = "dodge") +
  theme_bw() +
  theme(legend.position = "top") +
  scale_fill_brewer(palette = "Dark2") +
  ylab("Absolute frequencies")
```

And the second one:

```{r}
ggplot(data = titanic, aes(x = class, fill = survived)) +
  geom_bar(position = "fill") +
  theme_bw() +
  theme(legend.position = "top") +
  scale_fill_brewer(palette = "Dark2") +
  ylab("Relative frequencies")
```

## Summary exercise

The [contingencies]{.orange} are equal to the difference between observed frequencies
and expected frequencies, under the hypothesis of independence: $$
(\text{contingency}_{ij}) = n_{ij} - \hat{n}_{ij}, \qquad i=1,\dots,h,\quad j=1,\dots,k.
$$

Recall that, under independence, the [expected frequencies]{.blue} are
$\hat{n}_{ij} = n_{i+} n_{+j} / n$.

The $\chi^2$ index of association is defined as $$
\chi^2 = \sum_{i=1}^h\sum_{j=1}^k \frac{(n_{ij} - \hat{n}_{ij})^2}{\hat{n}_{ij}} = n\left(\sum_{i=1}^h\sum_{j=1}^k\frac{f_{ij}^2}{f_{i+}f_{+j}} - 1\right).
$$

Write an **R** function called `chi_squared(x, y)` that computes Pearson's
$\chi^2$ index.

#### Solution

```{r}
chi_squared <- function(x, y) {
  nn <- table(x, y)
  n <- sum(nn)
  ff <- nn / n # Joint relative frequencies
  f_x <- table(x) / n # Marginal relative frequencies of x
  f_y <- table(y) / n # Marginal relative frequencies of y
  S <- 0
  for (i in 1:length(f_x)) {
    for (j in 1:length(f_y)) {
      S <- S + ff[i, j]^2 / (f_x[i] * f_y[j])
    }
  }
  n * (S - 1)
}
chi_squared(titanic$survived, titanic$class)
```

#### Solution (alternative, more concise)

The following solution makes use of the functions `apply` and `outer`.

```{r}
chi_squared <- function(x, y) {
  nn <- table(x, y)
  n <- sum(nn)
  ff <- nn / n
  f_x <- apply(ff, 1, sum)
  f_y <- apply(ff, 2, sum)
  f_e <- outer(f_x, f_y) # "Outer" product between vectors
  n * (sum(ff^2 / f_e) - 1)
}
chi_squared(titanic$survived, titanic$class)
```

Finally, note that the function `chisq.test` produces the same result.

```{r}
chisq.test(table(titanic$survived, titanic$class))
```
