---
title: "Exercises A"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
lang: en
page-layout: full
bibliography: ../biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: false
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 200
    highlight-style: github
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type: console
---

## [Homepage](../index.html)

These exercises accompany [Unit A](../un_A.html). A few of them require
[writing your own functions]{.orange} or working with [data frames]{.blue},
which are covered later on: come back to those once we have seen the
corresponding material.

## Exercise A

Write in **R** the function `f(x)` associated with

$$
f(x) = \frac{|\sin(\pi/(4 x^2))|}{x}, \qquad x \neq 0.
$$

1.  What is the value of $f(1)$? And of $f(10^8)$?

2.  Approximate $f(1)$ and $f(10)$ to the first two decimal digits.

3.  Draw a plot of the function $f(x)$ over the interval $(1,3)$.

4.  Compute $\sum_{k=1}^{K} f(k)$, for $K = 500$.

5.  [Difficult]{.orange}. The `integrate` function of **R** (see the
    documentation) makes it possible to compute [numerically]{.blue}
    integrals of the form $$
    I = \int_a^b f(x)\mathrm{d}x,
    $$ where $a < b$ are two real values. Compute the value of the
    integral $\int_1^3 f(x)\mathrm{d}x$, where $f(x)$ is the function
    used in the previous points.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
f <- function(x) {
  abs(sin(pi / (4 * x^2))) / x
}

# A.1
f(1)
f(10^8)

# A.2
round(f(1), 2)
round(f(10), 2)

# A.3
curve(f(x), 1, 3)

# A.4
K <- 500
sum(f(1:K))

# A.5
integrate(f, lower = 1, upper = 3)
```
:::

## Exercise B

Verify numerically the following identities for $n = 10, 100, 1000$
(where possible).

1.  $$
    \sum_{k=0}^n(-1)^k\binom{n}{k} = 0.
    $$

2.  $$
    \sum_{k=1}^n k \binom{n}{k} = n 2^{n-1}.
    $$

3.  $$
    \sum_{k=1}^n(-1)^{k}k \binom{n}{k} = 0.
    $$

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# B.1
n <- 10
sum((-1)^(0:n) * choose(n, 0:n))

n <- 100
sum((-1)^(0:n) * choose(n, 0:n)) # NUMERICAL ERROR!

n <- 1000
sum((-1)^(0:n) * choose(n, 0:n)) # NUMERICAL ERROR!

# B.2
n <- 10
sum((1:n) * choose(n, 1:n))
n * 2^(n - 1)

n <- 100
sum((1:n) * choose(n, 1:n))
n * 2^(n - 1)

n <- 1000
sum((1:n) * choose(n, 1:n))
n * 2^(n - 1)

# B.3
n <- 10
sum((-1)^(1:n) * (1:n) * choose(n, 1:n))

n <- 100
sum((-1)^(1:n) * (1:n) * choose(n, 1:n)) # NUMERICAL ERROR!

n <- 1000
sum((-1)^(1:n) * (1:n) * choose(n, 1:n)) # NUMERICAL ERROR!
```
:::

## Exercise C

Write in **R** a function `roots(a, b, c)` that takes as input the
coefficients of the polynomial

$$
a x^2 + b x + c,
$$ and returns the vector of the values of $x$ that make it equal to
zero.

1.  Compute `roots(1, 5, 2)`;

2.  Compute `roots(1, 2, 1)`;

3.  Compute `roots(1, 1, 1)` and comment on the result.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# The solution of a quadratic equation is of course available in closed form

roots <- function(a, b, c) {
  Delta <- b^2 - 4 * a * c
  if (Delta < 0) stop("The discriminant must be non-negative")
  (-b + c(-1, 1) * sqrt(Delta)) / (2 * a)
}

# C.1
roots(1, 5, 2)

# C.2
roots(1, 2, 1)

# C.3
roots(1, 1, 1) # There are no real solutions
```
:::

## Exercise D

Define in **R** a symmetric matrix ${\bf A}$ of dimension $5 \times 5$,
whose elements are $$
{\bf A} = \begin{pmatrix}
26 &  22 &  17 &  22 &  23 \\ 
22 &  18 &  14 &  23 &  27 \\ 
17 &  14 &  14 &  20 &  24 \\ 
22 &  23 &  20 &  26 &  23 \\ 
23 &  27 &  24 &  23 &  12 \\ 
\end{pmatrix}.
$$

1.  Verify empirically that
    $\text{det}({\bf A}^{-1}) = 1 / \text{det}({\bf A})$.

2.  Verify empirically that
    $\text{det}({\bf A}^2) = \text{det}({\bf A})^2$.

3.  Compute the trace of ${\bf A}$, that is
    $\text{tr}({\bf A}) = \sum_{i=1}^n a_{ii}$, for $n = 5$.

4.  Verify empirically that
    $\text{det}({\bf A}) = \prod_{i=1}^n \lambda_i$, where
    $\lambda_1,\dots,\lambda_n$ are the eigenvalues of ${\bf A}$.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
A <- rbind(
  c(26, 22, 17, 22, 23),
  c(22, 18, 14, 23, 27),
  c(17, 14, 14, 20, 24),
  c(22, 23, 20, 26, 23),
  c(23, 27, 24, 23, 12)
)

# D.1
det(solve(A)) # 0.0006493506
1 / det(A) # Same value

# D.2
det(A %*% A) # 2371600
det(A)^2 # Same value

# D.3
sum(diag(A)) # 96

# D.4
det(A) # 1540
prod(eigen(A)$values) # Same value
```
:::

## Exercise E

The Euclidean distance between two vectors ${\bf x}$ and ${\bf y}$ in
$\mathbb{R}^p$ is defined as
$$d({\bf x},{\bf y}) = \sqrt{\sum_{j=1}^p(x_j - y_j)^2}.$$

1.  Write the **R** function `dist_euclid(x, y)` that computes the
    Euclidean distance between two vectors.

2.  If `x = c(1, 4, 2, 2, 10)` and `y = c(8, 1, 8, 3, 6)`, what is the
    value of `dist_euclid(x, y)`?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# E.1
dist_euclid <- function(x, y) {
  sqrt(crossprod(x - y))
}

# E.2
x <- c(1, 4, 2, 2, 10)
y <- c(8, 1, 8, 3, 6)
dist_euclid(x, y) # 10.53565
```
:::

## Exercise F

The Frobenius norm $||{\bf A}||_F$ of a real-valued matrix ${\bf A}$ of
dimension $n \times m$ is defined as
$$||{\bf A}||_F = \sqrt{\sum_{i=1}^n \sum_{j=1}^m a_{ij}^2}.$$

1.  Write the **R** function `frobenius_norm(A)` that computes the
    Frobenius norm of a matrix `A`.

2.  If `A = cbind(1:5,6:10,11:15)`, what is the value of
    `frobenius_norm(A)`?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# F.1
frobenius_norm <- function(A) {
  sqrt(sum(A^2))
}

# F.2
A <- cbind(1:5, 6:10, 11:15)
frobenius_norm(A) # 35.21363
```
:::

## Exercise G

Consider the dataset `data(mtcars)`, which is available in **R**. Answer
the following questions:

1.  How many rows and how many columns does this dataset have?

2.  Why does the command `mtcars[1:20]` return an error? What is the
    difference with the command `mtcars[1:20, ]`?

3.  Each of the following lines contains an error (a bug). Propose a
    modification to fix it

```{r}
#| eval: false
mtcars[mtcars$cyl = 4, ] 
mtcars[-1:4, ] 
mtcars[mtcars$cyl <= 5] 
mtcars[mtcars$cyl == 4 | 6, ]
```

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
data(mtcars)

# G.1
dim(mtcars) # 32 rows, 11 columns

# G.2
mtcars[1:20] # A comma is missing
mtcars[1:20, ] # Correct command

# G.3
# mtcars[mtcars$cyl = 4, ] # WRONG
mtcars[mtcars$cyl == 4, ] # CORRECT

# mtcars[-1:4, ] # WRONG
mtcars[-(1:4), ] # CORRECT (drops the first 4 rows)

# mtcars[mtcars$cyl <= 5] # WRONG
mtcars[mtcars$cyl <= 5, ] # CORRECT

# mtcars[mtcars$cyl == 4 | 6, ] # WRONG
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ] # CORRECT
```
:::

## Exercise H

Let ${\bf A}$ be a matrix of dimension $n \times n$. Implement a function
that extracts the diagonal elements $a_{ii}$, for $i = 1,\dots,n$, of the
matrix ${\bf A}$. Then compare it with the `diag` function of **R**.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
diag2 <- function(A) {
  n <- nrow(A)
  diagonal <- numeric(n)
  for (i in 1:n) {
    diagonal[i] <- A[i, i]
  }
  diagonal
}

A <- rbind(
  c(26, 22, 17, 22, 23),
  c(22, 18, 14, 23, 27),
  c(17, 14, 14, 20, 24),
  c(22, 23, 20, 26, 23),
  c(23, 27, 24, 23, 12)
)

diag(A)
diag2(A)
```
:::

## Exercise I

What does the function `f(10)` return? Discuss the result without running
the following code

```{r}
#| eval: false
h <- function(x) {
  x^2 
}

g <- function(x) {
  h(x) + 1 
}

f <- function(x) {
  g(x) * 2 
}

f(10)
```

::: {.callout-note collapse="true"}
#### Solution

The exercise asked you to guess the result without running the code. If
executed, it returns `202`.
:::
