---
title: "Exercises B"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
lang: en
page-layout: full
bibliography: ../biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: false
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 200
    highlight-style: github
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type: console
---

## [Homepage](../index.html)

These exercises accompany Units [D](../un_D.html), [E](../un_E.html),
[F](../un_F.html) and [G](../un_G.html).

The exercises whose title carries a number such as [(**AK** 1.10)]{.grey} are
taken from Chapter 1 of @Agresti2022; their data are those of the [website of
the book](https://stat4ds.rwth-aachen.de/data/).

## Exercise A

The sample variance of the data ${\bf x} = (x_1,\dots,x_n)$ is defined as

$$
\text{var}({\bf x}) = \frac{1}{n}\sum_{i=1}^n(x_i - \bar{x})^2,
$$ where $\bar{x}$ is the sample mean. Note that $\text{var}({\bf x})$ admits
the alternative representation

$$\text{var}({\bf x}) = \frac{1}{2 n^2} \sum_{i=1}^n \sum_{j = 1}^n (x_i - x_j)^2.$$

------------------------------------------------------------------------

1.  Write a function `var2(x)` that computes the variance of ${\bf x}$ using the
    definition.

2.  Write a function `var3(x)` that computes the variance of ${\bf x}$ using the
    formula based on the distances between pairs of elements.

3.  Suppose that `x = c(1, 4, 2, 2, 10)`. Check that the two functions `var2(x)`
    and `var3(x)` return the same result.

4.  Suppose now that `x <- 1:3000`. Do you notice any difference with respect to
    the previous point?

5.  Compare the functions `var2` and `var3` with the function `var` implemented
    in **R**, using the data of point 3. Why do the results differ, even if only
    slightly? Consult the documentation to answer.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# A.1
var2 <- function(x) {
  mean((x - mean(x))^2)
}

# A.2
var3 <- function(x) {
  n <- length(x) # Get the length of the vector x
  D <- matrix(0, nrow = n, ncol = n) # Creation of an empty matrix
  for (i in 1:n) {
    for (j in 1:n) {
      D[i, j] <- (x[i] - x[j])^2
    }
  }
  sum(D) / (2 * n^2)
}

# A.3
x <- c(1, 4, 2, 2, 10)
c(var2(x), var3(x))

# A.4
x <- 1:3000
var2(x)
var3(x) # Much slower than var2(x)

# A.5
var(x) # It divides by n - 1 and not by n
```
:::

## Exercise B

At [this link](../data/elections.RData) the data of the 2016 municipal elections
in the city of Milan are available. Each statistical unit is a single vote; the
variable `party` reports the list the vote was cast for, and the variable
`district` the borough (*Municipio*) in which it was cast.

------------------------------------------------------------------------

1.  Load the data into memory.

2.  Which variables does the dataset `elections` contain? Of which type are
    they?

3.  Obtain the number of votes obtained by each party in the different districts
    of Milan.

4.  Obtain the number of votes obtained by each party and draw a plot of it.
    Obtain moreover the percentage of votes obtained by each party.

5.  Write the function `Gini(x)` that computes the Gini index for a qualitative
    variable. The Gini index is defined as $$
    G = 1 - \sum_{j=1}^k f_j^2,
    $$ where $f_1,\dots,f_k$ are the relative frequencies of the $k$ categories.

6.  Write a function `Gini_norm(x)` that computes the normalised Gini index,
    that is $$G_\text{norm} = \frac{k}{k-1} G.$$

7.  Evaluate the fragmentation of the votes among the parties in the different
    districts using the normalised Gini index. Organise the results in a table
    and say which district is the most fragmented and which one is the most
    concentrated, that is the one in which the votes are polarised towards a few
    parties. Pay attention to the direction of the index: a Gini index close to
    $1$ means that the votes are spread evenly across the parties.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# B.1
# Assuming the file has been downloaded and the "path" is the correct one:
load("data/elections.RData")

# B.2
str(elections) # Two variables of type "factor", "party" and "district"

# B.3
table(elections$party, elections$district)

# B.4
# Number of votes
table(elections$party)

# Barplot (the names may not appear, depending on the size of the plot window)
barplot(table(elections$party))

# Percentage of votes (rounded)
round(100 * table(elections$party) / nrow(elections), 2)

# B.5
Gini <- function(x) {
  freq <- as.numeric(table(x))
  freq_rel <- freq / sum(freq)
  1 - sum(freq_rel^2)
}

# B.6
Gini_norm <- function(x) {
  freq <- as.numeric(table(x))
  freq_rel <- freq / sum(freq)
  k <- length(freq)
  k / (k - 1) * (1 - sum(freq_rel^2))
}

# B.7
G <- tapply(elections$party, elections$district, Gini_norm)
round(G, 4)
names(which.max(G)) # District 9: the most fragmented
names(which.min(G)) # District 1: the most concentrated (polarised)
```
:::

## Exercise C

The data available at [this link](../data/swiss.csv) refer to $47$
French-speaking provinces of Switzerland around 1888. For each province three
variables are reported: `fertility`, a standardised fertility measure;
`agriculture`, the percentage of males employed in agriculture; and `education`,
the logarithm of the percentage of people with education beyond primary school.

------------------------------------------------------------------------

1.  Load the data into memory and save the dataset in the object `swiss_data`.

2.  Which variables does the dataset contain? Of which type are they?

3.  Represent the variable `education` graphically through a histogram. Then
    compute the mean, the median and the variance.

4.  Represent graphically the cumulative distribution function of the variable
    `agriculture`.

5.  Draw the scatterplot of the variables `agriculture` and `education`. Can you
    notice any relationship?

6.  Obtain the variance-covariance matrix.

7.  Obtain the correlation matrix. Comment on the relationship between
    `agriculture` and `education`.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# C.1
swiss_data <- read.table("data/swiss.csv", header = TRUE)

# C.2
str(swiss_data) # Three numerical variables

# C.3
hist(swiss_data$education)
summary(swiss_data$education)
var(swiss_data$education) # Leaving aside the n vs n-1 denominator

# C.4
plot(ecdf(swiss_data$agriculture))

# C.5
plot(swiss_data$agriculture, swiss_data$education)

# C.6
cov(swiss_data) # Leaving aside the n vs n-1 denominator

# C.7
cor(swiss_data)
# Between agriculture and education there is a fairly marked negative relationship
```
:::

## Exercise D

In the dataset `football` available at [this link](../data/football.csv), each
statistical unit represents a football match of the Italian Serie A, for the
2008-2015 championships. The dataset is made of the following variables:

-   `Date` indicates the date of the match.

-   `HomeTeam` and `AwayTeam` indicate, respectively, the team playing at home
    and the visiting team for each match.

-   `FTR` indicates whether the team playing at home won (H), drew (D), or lost
    (A).

-   The variables `B365H`, `B365D`, `B365A` are the odds for the victory, draw
    and defeat of the home team.

Supposing to bet on the victory of the home team, betting one euro on it one
would obtain `B365H` euros in case of victory.

------------------------------------------------------------------------

1.  Load the dataset into memory.

2.  Check the types of the variables present in the dataset `football`.

3.  Check whether there are missing data in the dataset. If present, exclude the
    rows containing missing values from the analysis. **Hint**: use the function
    `na.omit`.

4.  Compute the main descriptive indices for the variable `B365H`. Then
    represent it through a histogram and a boxplot. Are there outliers? How can
    one tell?

5.  Carry out the analysis of the previous point on the logarithmic
    transformation of `B365H`. Are there outliers?

6.  Evaluate the correlation between `B365H` and `B365A`. Could one have
    expected a similar result? What would a positive correlation imply?

7.  Represent with a suitable plot the relationship between the logarithmic
    transformations of the variables `B365H` and `B365A`. Comment on the result.

8.  Evaluate the average odds `B365H` for each category of the variable `FTR`.
    How is this result interpreted?

9.  Represent through a boxplot the variable `B365H` for each category of the
    variable `FTR`. Comment on the result.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# D.1
football <- read.table("data/football.csv",
  header = TRUE, sep = ",", stringsAsFactors = TRUE
)

# D.2
str(football)

# D.3
summary(football) # There are 3 missing values
football <- na.omit(football) # Missing data are excluded and the dataset overwritten

# D.4
summary(football$B365H)

par(mfrow = c(1, 2))
hist(football$B365H, breaks = 20)
boxplot(football$B365H) # The plot shows that there are several outliers

# D.5
summary(log(football$B365H))
par(mfrow = c(1, 2))
hist(log(football$B365H), breaks = 20)
boxplot(log(football$B365H)) # Outliers are still present, but the effect is reduced

# D.6
cor(football$B365H, football$B365A) # The comment is left to the student

# D.7
par(mfrow = c(1, 1))
plot(log(football$B365H), log(football$B365A)) # The comment is left to the student

# D.8
tapply(football$B365H, football$FTR, mean) # The comment is left to the student

# D.9
plot(football$B365H ~ football$FTR) # The comment is left to the student
```
:::

## Exercise E

In the dataset `imdb` available at [this link](../data/imdb.RData), each
statistical unit is a film extracted from the website http://www.imdb.com/
(Internet Movie Data Base). For each film, the following variables are
available:

-   `movie_title` is the title of the film.

-   `duration` is the duration of the film expressed in minutes.

-   `lgross` is the logarithmic transformation of the gross revenue, expressed
    in dollars.

-   `lbudget` is the logarithmic transformation of the budget, expressed in
    dollars.

-   `Action, Adventure, Animation, Comedy, Crime, Documentary, Drama, Family, Fantasy, History, Horror, Music, Romance, SciFi, Thriller, War, Western`
    are qualitative variables indicating whether the film belongs to the genre
    or not. Each film can belong to several genres at the same time.

------------------------------------------------------------------------

1.  Load the dataset into memory. Check the type of the variables.

2.  How many statistical units are there?

3.  What is the average duration of the films in the dataset? And the average
    gross revenue?

4.  Are there differences between the distribution of the duration (`duration`)
    of dramatic films (`Drama`) and non-dramatic ones? Answer through suitable
    descriptive indices and graphical analyses.

5.  Which are the 5 films with the highest gross revenue within the dataset?
    **Hint**: use the function `order`.

6.  Represent with appropriate plots the marginal distributions and the joint
    distribution of the variables `lbudget` and `lgross`. What kind of
    relationship seems to be there between the two variables?

7.  Compute the correlation between `lbudget` and `lgross` and comment on the
    result.

8.  Compute the correlation between `duration` and `lgross` and comment on the
    result. Can we therefore conclude that, in order to earn a lot, it is enough
    to produce a film lasting 12 hours?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# E.1
load("data/imdb.RData")
str(imdb)

# E.2
nrow(imdb)

# E.3
mean(imdb$duration)
mean(exp(imdb$lgross))

# E.4
tapply(imdb$duration, imdb$Drama, mean)
tapply(imdb$duration, imdb$Drama, median)
plot(imdb$duration ~ imdb$Drama) # Boxplot

# E.5
imdb$movie_title[order(imdb$lgross, decreasing = TRUE)][1:5]

# E.6
par(mfrow = c(1, 3))
hist(imdb$lbudget)
hist(imdb$lgross)
plot(imdb$lbudget, imdb$lgross)
par(mfrow = c(1, 1))

# E.7
cor(imdb$lbudget, imdb$lgross) # The comment is left to the student

# E.8
cor(imdb$duration, imdb$lgross) # The comment is left to the student
```
:::

## Exercise F [(**AK** 1.10)]{.grey}

The `Carbon_West` data file of @Agresti2022, available at [this
link](https://stat4ds.rwth-aachen.de/data/Carbon_West.dat), reports the [per
capita CO$_2$ emissions]{.orange} (in metric tons) of $35$ Western nations.

------------------------------------------------------------------------

1.  Import the data into **R**. The `read.table` function accepts a [web
    address]{.orange} in place of a file name, so there is no need to download
    the file first.

2.  Construct a frequency distribution and a histogram of `CO2`. How would you
    describe the shape of the distribution?

3.  Find the mean, the median and the standard deviation, and interpret them.

4.  Which nations are responsible for the right tail of the distribution?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# F.1
path <- "https://stat4ds.rwth-aachen.de/data/Carbon_West.dat"
Carbon <- read.table(path, header = TRUE)
str(Carbon) # 35 nations, a single quantitative variable

# F.2
classes <- seq(0, 18, by = 3)
table(cut(Carbon$CO2, breaks = classes))
hist(Carbon$CO2, breaks = classes)
# Most nations emit between 3 and 9 tons, with a long right tail

# F.3
mean(Carbon$CO2) # 6.72
median(Carbon$CO2) # 5.90, smaller than the mean, as expected for a right tail
sd(Carbon$CO2) # 3.36, leaving aside the n vs n - 1 denominator

# F.4
Carbon$Nation[order(Carbon$CO2, decreasing = TRUE)][1:3]
# The United States, Australia and Canada
```
:::

## Exercise G [(**AK** 1.17)]{.grey}

The `Murder` data file of @Agresti2022, available at [this
link](https://stat4ds.rwth-aachen.de/data/Murder.dat), reports the [murder
rate]{.orange} (number of murders per $100{,}000$ inhabitants) of each of the
$50$ states of the U.S. and of the District of Columbia (`DC`), according to
the FBI Uniform Crime Reports.

At first, [leave the observation relative to `DC` aside]{.blue}.

------------------------------------------------------------------------

1.  Import the data into **R** and create a second data frame, without the row
    relative to `DC`.

2.  Using the data without `DC`, find the mean and the standard deviation, and
    interpret their values.

3.  Still without `DC`, find the five-number summary and construct the
    corresponding boxplot. Interpret them.

4.  Include `DC` again. What is affected more by this outlier: the mean or the
    median? The range or the interquartile range? Why?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# G.1
path <- "https://stat4ds.rwth-aachen.de/data/Murder.dat"
Murder <- read.table(path, header = TRUE)
Murder_no_DC <- subset(Murder, state != "DC")

# G.2
mean(Murder_no_DC$murder) # 4.87 murders per 100.000 inhabitants
sd(Murder_no_DC$murder) # 2.59

# G.3
fivenum(Murder_no_DC$murder) # 1.00, 2.60, 4.85, 6.20, 12.40
boxplot(Murder_no_DC$murder)
# Louisiana (12.4) is already an outlier among the 50 states

# G.4
c(mean(Murder_no_DC$murder), mean(Murder$murder)) # 4.87 -> 5.25
c(median(Murder_no_DC$murder), median(Murder$murder)) # 4.85 -> 5.00

diff(range(Murder_no_DC$murder)) # 11.4
diff(range(Murder$murder)) # 23.2, twice as much

IQR(Murder_no_DC$murder) # 3.55
IQR(Murder$murder) # 3.80, almost unchanged

# A single observation moves the mean and the range a lot, the median and the
# interquartile range very little: the latter are *robust* to outliers
```
:::

## Exercise H [(**AK** 1.19)]{.grey}

The `Houses` data file of @Agresti2022, available at [this
link](https://stat4ds.rwth-aachen.de/data/Houses.dat), lists the selling price
(`price`, in thousands of dollars), the size (`size`, in square feet), the tax
bill (`taxes`, in dollars), the number of bedrooms (`bedrooms`) and of
bathrooms (`baths`), and whether the house is new (`new`, $1 =$ yes, $0 =$ no)
for $100$ home sales in Gainesville, Florida. Here we analyse the [selling
prices]{.orange}.

------------------------------------------------------------------------

1.  Import the data into **R**. Construct a frequency distribution and a
    histogram of `price`, and describe the shape of the distribution.

2.  Find the percentage of observations that fall within one standard deviation
    of the mean, that is in the interval $(\bar{x} - s, \bar{x} + s)$. Why is
    this percentage not close to $68\%$?

3.  Construct a boxplot and interpret it.

4.  Use descriptive statistics to compare the selling prices of new and old
    houses. Would you regard this comparison as reliable?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# H.1
path <- "https://stat4ds.rwth-aachen.de/data/Houses.dat"
Houses <- read.table(path, header = TRUE)

classes <- seq(0, 900, by = 100)
table(cut(Houses$price, breaks = classes))
hist(Houses$price, breaks = classes)
# The distribution is strongly skewed to the right

# H.2
price_mean <- mean(Houses$price) # 233.0
price_sd <- sd(Houses$price) # 151.9
mean(abs(Houses$price - price_mean) <= price_sd) # 0.85, that is 85%
# The empirical rule concerns bell-shaped distributions. Here the distribution
# is strongly asymmetric, so the rule does not apply

# H.3
fivenum(Houses$price) # 31.5, 138.8, 198.9, 257.6, 880.5
boxplot(Houses$price) # Several outliers, all on the right

# H.4
table(Houses$new) # Only 11 of the 100 houses are new
tapply(Houses$price, Houses$new, mean) # 207.9 vs 436.4
tapply(Houses$price, Houses$new, median) # 190.8 vs 427.5
boxplot(price ~ new, data = Houses)
# New houses are sold at about twice the price, but they are only 11: the
# comparison is based on a very small group
```
:::

## Exercise I [(**AK** 1.20)]{.grey}

Refer to the previous exercise, and let $y =$ `price` and $x =$ `size`.

------------------------------------------------------------------------

1.  Construct a scatterplot and interpret it. Identify the observation that
    seems to fall apart from the others.

2.  Find the correlation and interpret it.

3.  Find the regression line. Interpret the slope, and find the predicted
    selling price for a house of (i) $1000$ square feet, (ii) $4000$ square
    feet.

4.  Remove the observation identified at point 1 and obtain the regression line
    again. How much do the estimates change?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# I.1
plot(Houses$size, Houses$price)
# One house is the largest of the sample but is sold at an average price
Houses[which.max(Houses$size), ] # Case 64: 4050 square feet, 337.5

# I.2
cor(Houses$size, Houses$price) # 0.834, a strong positive linear relationship

# I.3
fit <- lm(price ~ size, data = Houses)
coef(fit) # Intercept -76.39, slope 0.1899
# One additional square foot is worth, on average, about 190 dollars
abline(fit)

predict(fit, newdata = data.frame(size = c(1000, 4000))) # 113.5 and 683.2

# I.4
Houses_no64 <- subset(Houses, case != 64)
fit2 <- lm(price ~ size, data = Houses_no64)
coef(fit2) # Intercept -109.37, slope 0.2127
cor(Houses_no64$size, Houses_no64$price) # 0.871
# A single observation out of 100 changes the slope by more than 10%
```
:::

## Exercise J [(**AK** 1.24)]{.grey}

The `UN` data file of @Agresti2022, available at [this
link](https://stat4ds.rwth-aachen.de/data/UN.dat), has United Nations data for
$42$ nations on the following variables:

-   `GDP`, the per capita gross domestic product, in thousands of dollars;

-   `HDI`, a human development index, with components referring to life
    expectancy at birth, educational attainment and income per capita;

-   `GII`, a gender inequality index, a composite measure reflecting inequality
    in achievement between women and men in reproductive health, empowerment
    and the labour market;

-   `Fertility`, the fertility rate, that is the number of births per woman;

-   `CO2`, the [carbon dioxide emissions per capita]{.orange}, in metric tons;

-   `Homicide`, the homicide rate, per $100{,}000$ people;

-   `Prison`, the prison population, per $100{,}000$ people;

-   `Internet`, the percentage of people using the Internet.

------------------------------------------------------------------------

1.  Import the data into **R** and conduct a descriptive analysis of the prison
    rates, highlighting any unusual observation.

2.  Construct the correlation matrix of the eight quantitative variables. Which
    pair has the strongest association?

3.  Study the association between `CO2` and `GDP` through a scatterplot, the
    correlation and the regression line. Show how the predicted CO$_2$ varies
    as `GDP` goes from its minimum to its maximum value in this sample.

4.  The correlation between `GDP` and `Prison` is essentially zero. Does this
    mean that the two variables are unrelated? Look at the scatterplot, and
    then repeat the computation without the United States.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# J.1
path <- "https://stat4ds.rwth-aachen.de/data/UN.dat"
UN <- read.table(path, header = TRUE)

summary(UN$Prison) # Median 119.5, mean 153.9: the distribution is skewed
sd(UN$Prison) # 124.3
boxplot(UN$Prison)
UN$Nation[order(UN$Prison, decreasing = TRUE)][1:3]
# The United States (716) are far above every other nation, Russia is second

# J.2
round(cor(UN[, -1]), 2) # The first column contains the names of the nations
# The strongest association is between GDP and HDI (0.90), which is no
# surprise, since the income per capita is one of the components of the HDI

# J.3
plot(UN$GDP, UN$CO2)
cor(UN$GDP, UN$CO2) # 0.674

fit <- lm(CO2 ~ GDP, data = UN)
coef(fit) # Intercept 1.926, slope 0.1778
abline(fit)

predict(fit, newdata = data.frame(GDP = range(UN$GDP))) # From 2.71 to 13.11

# J.4
cor(UN$GDP, UN$Prison) # -0.003
plot(UN$GDP, UN$Prison) # The United States are an extreme observation

UN_no_US <- subset(UN, Nation != "US")
cor(UN_no_US$GDP, UN_no_US$Prison) # -0.27
# A correlation close to zero does not mean that there is no relationship: here
# it is the result of a single, very influential observation
```
:::

## Exercise K [(**AK** 1.27)]{.grey}

A study of sheep analysed whether a sheep survived for a year from the original
observation time (`survival`, $1 =$ yes, $0 =$ no) as a function of its
`weight`, in kg, at the original observation. The `Sheep` data file of
@Agresti2022 is available at [this
link](https://stat4ds.rwth-aachen.de/data/Sheep.dat).

------------------------------------------------------------------------

1.  Import the data into **R**. How many sheep are there? How many of them
    survived?

2.  Compare the weights of the sheep that survived with the weights of the
    sheep that did not, using numerical descriptive statistics.

3.  Compare the same two distributions graphically.

4.  Summarise the results of your analysis in a few sentences. Can you conclude
    that being heavier [causes]{.orange} a sheep to survive?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# K.1
path <- "https://stat4ds.rwth-aachen.de/data/Sheep.dat"
Sheep <- read.table(path, header = TRUE)

nrow(Sheep) # 1359 sheep
table(Sheep$survival) # 318 did not survive, 1041 did

# K.2
tapply(Sheep$weight, Sheep$survival, mean) # 16.0 vs 20.6
tapply(Sheep$weight, Sheep$survival, median) # 14.8 vs 21.6
tapply(Sheep$weight, Sheep$survival, sd) # 5.3 vs 4.9
tapply(Sheep$weight, Sheep$survival, summary)

# K.3
boxplot(weight ~ survival, data = Sheep)

# K.4
# The sheep that survived are clearly heavier, by about 5 kg on average, and
# the two distributions still overlap a great deal. This is an observational
# study: the weight is a symptom of the health of the animal, not necessarily
# a cause of its survival
```
:::
