---
title: "Exercises B"
subtitle: "Programming and Data Analysis with R"
author: "[Tommaso Rigon]{.orange}"
institute: "_Università degli Studi di Milano-Bicocca_"
lang: en
page-layout: full
bibliography: ../biblio.bib
citeproc: true
csl: https://www.zotero.org/styles/journal-of-the-american-statistical-association
reference-location: margin
execute:
  cache: false
format:
  html:
    html-math-method: katex
    echo: true
    callout-appearance: minimal
    theme: [simplex, ../template.css]
    toc: true
    toc-title: Table of contents
    embed-resources: false
    code-line-numbers: false
    smooth-scroll: true
    code-fold: false
    code-summary: "Show the code"
    fig-dpi: 200
    highlight-style: github
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type: console
---

## [Homepage](../index.html)

These exercises accompany Units [D](../un_D.html), [E](../un_E.html),
[F](../un_F.html) and [G](../un_G.html).

## Exercise A

The sample variance of the data ${\bf x} = (x_1,\dots,x_n)$ is defined as

$$
\text{var}({\bf x}) = \frac{1}{n}\sum_{i=1}^n(x_i - \bar{x})^2,
$$ where $\bar{x}$ is the sample mean. Note that $\text{var}({\bf x})$ admits
the alternative representation

$$\text{var}({\bf x}) = \frac{1}{2 n^2} \sum_{i=1}^n \sum_{j = 1}^n (x_i - x_j)^2.$$

------------------------------------------------------------------------

1.  Write a function `var2(x)` that computes the variance of ${\bf x}$ using the
    definition.

2.  Write a function `var3(x)` that computes the variance of ${\bf x}$ using the
    formula based on the distances between pairs of elements.

3.  Suppose that `x = c(1, 4, 2, 2, 10)`. Check that the two functions `var2(x)`
    and `var3(x)` return the same result.

4.  Suppose now that `x <- 1:3000`. Do you notice any difference with respect to
    the previous point?

5.  Compare the functions `var2` and `var3` with the function `var` implemented
    in **R**, using the data of point 3. Why do the results differ, even if only
    slightly? Consult the documentation to answer.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# A.1
var2 <- function(x) {
  mean((x - mean(x))^2)
}

# A.2
var3 <- function(x) {
  n <- length(x) # Get the length of the vector x
  D <- matrix(0, nrow = n, ncol = n) # Creation of an empty matrix
  for (i in 1:n) {
    for (j in 1:n) {
      D[i, j] <- (x[i] - x[j])^2
    }
  }
  sum(D) / (2 * n^2)
}

# A.3
x <- c(1, 4, 2, 2, 10)
c(var2(x), var3(x))

# A.4
x <- 1:3000
var2(x)
var3(x) # Much slower than var2(x)

# A.5
var(x) # It divides by n - 1 and not by n
```
:::

## Exercise B

At [this link](../data/elections.RData) the data of the 2016 municipal elections
in the city of Milan are available. Each statistical unit is a single vote; the
variable `party` reports the list the vote was cast for, and the variable
`district` the borough (*Municipio*) in which it was cast.

------------------------------------------------------------------------

1.  Load the data into memory.

2.  Which variables does the dataset `elections` contain? Of which type are
    they?

3.  Obtain the number of votes obtained by each party in the different districts
    of Milan.

4.  Obtain the number of votes obtained by each party and draw a plot of it.
    Obtain moreover the percentage of votes obtained by each party.

5.  Write the function `Gini(x)` that computes the Gini index for a qualitative
    variable. The Gini index is defined as $$
    G = 1 - \sum_{j=1}^k f_j^2,
    $$ where $f_1,\dots,f_k$ are the relative frequencies of the $k$ categories.

6.  Write a function `Gini_norm(x)` that computes the normalised Gini index,
    that is $$G_\text{norm} = \frac{k}{k-1} G.$$

7.  Evaluate the fragmentation of the votes among the parties in the different
    districts using the normalised Gini index. Organise the results in a table
    and say which district is the most fragmented and which one is the most
    concentrated, that is the one in which the votes are polarised towards a few
    parties. Pay attention to the direction of the index: a Gini index close to
    $1$ means that the votes are spread evenly across the parties.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# B.1
# Assuming the file has been downloaded and the "path" is the correct one:
load("data/elections.RData")

# B.2
str(elections) # Two variables of type "factor", "party" and "district"

# B.3
table(elections$party, elections$district)

# B.4
# Number of votes
table(elections$party)

# Barplot (the names may not appear, depending on the size of the plot window)
barplot(table(elections$party))

# Percentage of votes (rounded)
round(100 * table(elections$party) / nrow(elections), 2)

# B.5
Gini <- function(x) {
  freq <- as.numeric(table(x))
  freq_rel <- freq / sum(freq)
  1 - sum(freq_rel^2)
}

# B.6
Gini_norm <- function(x) {
  freq <- as.numeric(table(x))
  freq_rel <- freq / sum(freq)
  k <- length(freq)
  k / (k - 1) * (1 - sum(freq_rel^2))
}

# B.7
G <- tapply(elections$party, elections$district, Gini_norm)
round(G, 4)
names(which.max(G)) # District 9: the most fragmented
names(which.min(G)) # District 1: the most concentrated (polarised)
```
:::

## Exercise C

The data available at [this link](../data/swiss.csv) refer to $47$
French-speaking provinces of Switzerland around 1888. For each province three
variables are reported: `fertility`, a standardised fertility measure;
`agriculture`, the percentage of males employed in agriculture; and `education`,
the logarithm of the percentage of people with education beyond primary school.

------------------------------------------------------------------------

1.  Load the data into memory and save the dataset in the object `swiss_data`.

2.  Which variables does the dataset contain? Of which type are they?

3.  Represent the variable `education` graphically through a histogram. Then
    compute the mean, the median and the variance.

4.  Represent graphically the cumulative distribution function of the variable
    `agriculture`.

5.  Draw the scatterplot of the variables `agriculture` and `education`. Can you
    notice any relationship?

6.  Obtain the variance-covariance matrix.

7.  Obtain the correlation matrix. Comment on the relationship between
    `agriculture` and `education`.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# C.1
swiss_data <- read.table("data/swiss.csv", header = TRUE)

# C.2
str(swiss_data) # Three numerical variables

# C.3
hist(swiss_data$education)
summary(swiss_data$education)
var(swiss_data$education) # Leaving aside the n vs n-1 denominator

# C.4
plot(ecdf(swiss_data$agriculture))

# C.5
plot(swiss_data$agriculture, swiss_data$education)

# C.6
cov(swiss_data) # Leaving aside the n vs n-1 denominator

# C.7
cor(swiss_data)
# Between agriculture and education there is a fairly marked negative relationship
```
:::

## Exercise D

In the dataset `football` available at [this link](../data/football.csv), each
statistical unit represents a football match of the Italian Serie A, for the
2008-2015 championships. The dataset is made of the following variables:

-   `Date` indicates the date of the match.

-   `HomeTeam` and `AwayTeam` indicate, respectively, the team playing at home
    and the visiting team for each match.

-   `FTR` indicates whether the team playing at home won (H), drew (D), or lost
    (A).

-   The variables `B365H`, `B365D`, `B365A` are the odds for the victory, draw
    and defeat of the home team.

Supposing to bet on the victory of the home team, betting one euro on it one
would obtain `B365H` euros in case of victory.

------------------------------------------------------------------------

1.  Load the dataset into memory.

2.  Check the types of the variables present in the dataset `football`.

3.  Check whether there are missing data in the dataset. If present, exclude the
    rows containing missing values from the analysis. **Hint**: use the function
    `na.omit`.

4.  Compute the main descriptive indices for the variable `B365H`. Then
    represent it through a histogram and a boxplot. Are there outliers? How can
    one tell?

5.  Carry out the analysis of the previous point on the logarithmic
    transformation of `B365H`. Are there outliers?

6.  Evaluate the correlation between `B365H` and `B365A`. Could one have
    expected a similar result? What would a positive correlation imply?

7.  Represent with a suitable plot the relationship between the logarithmic
    transformations of the variables `B365H` and `B365A`. Comment on the result.

8.  Evaluate the average odds `B365H` for each category of the variable `FTR`.
    How is this result interpreted?

9.  Represent through a boxplot the variable `B365H` for each category of the
    variable `FTR`. Comment on the result.

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# D.1
football <- read.table("data/football.csv",
  header = TRUE, sep = ",", stringsAsFactors = TRUE
)

# D.2
str(football)

# D.3
summary(football) # There are 3 missing values
football <- na.omit(football) # Missing data are excluded and the dataset overwritten

# D.4
summary(football$B365H)

par(mfrow = c(1, 2))
hist(football$B365H, breaks = 20)
boxplot(football$B365H) # The plot shows that there are several outliers

# D.5
summary(log(football$B365H))
par(mfrow = c(1, 2))
hist(log(football$B365H), breaks = 20)
boxplot(log(football$B365H)) # Outliers are still present, but the effect is reduced

# D.6
cor(football$B365H, football$B365A) # The comment is left to the student

# D.7
par(mfrow = c(1, 1))
plot(log(football$B365H), log(football$B365A)) # The comment is left to the student

# D.8
tapply(football$B365H, football$FTR, mean) # The comment is left to the student

# D.9
plot(football$B365H ~ football$FTR) # The comment is left to the student
```
:::

## Exercise E

In the dataset `imdb` available at [this link](../data/imdb.RData), each
statistical unit is a film extracted from the website http://www.imdb.com/
(Internet Movie Data Base). For each film, the following variables are
available:

-   `movie_title` is the title of the film.

-   `duration` is the duration of the film expressed in minutes.

-   `lgross` is the logarithmic transformation of the gross revenue, expressed
    in dollars.

-   `lbudget` is the logarithmic transformation of the budget, expressed in
    dollars.

-   `Action, Adventure, Animation, Comedy, Crime, Documentary, Drama, Family, Fantasy, History, Horror, Music, Romance, SciFi, Thriller, War, Western`
    are qualitative variables indicating whether the film belongs to the genre
    or not. Each film can belong to several genres at the same time.

------------------------------------------------------------------------

1.  Load the dataset into memory. Check the type of the variables.

2.  How many statistical units are there?

3.  What is the average duration of the films in the dataset? And the average
    gross revenue?

4.  Are there differences between the distribution of the duration (`duration`)
    of dramatic films (`Drama`) and non-dramatic ones? Answer through suitable
    descriptive indices and graphical analyses.

5.  Which are the 5 films with the highest gross revenue within the dataset?
    **Hint**: use the function `order`.

6.  Represent with appropriate plots the marginal distributions and the joint
    distribution of the variables `lbudget` and `lgross`. What kind of
    relationship seems to be there between the two variables?

7.  Compute the correlation between `lbudget` and `lgross` and comment on the
    result.

8.  Compute the correlation between `duration` and `lgross` and comment on the
    result. Can we therefore conclude that, in order to earn a lot, it is enough
    to produce a film lasting 12 hours?

::: {.callout-note collapse="true"}
#### Solution

```{r}
#| eval: false
# E.1
load("data/imdb.RData")
str(imdb)

# E.2
nrow(imdb)

# E.3
mean(imdb$duration)
mean(exp(imdb$lgross))

# E.4
tapply(imdb$duration, imdb$Drama, mean)
tapply(imdb$duration, imdb$Drama, median)
plot(imdb$duration ~ imdb$Drama) # Boxplot

# E.5
imdb$movie_title[order(imdb$lgross, decreasing = TRUE)][1:5]

# E.6
par(mfrow = c(1, 3))
hist(imdb$lbudget)
hist(imdb$lgross)
plot(imdb$lbudget, imdb$lgross)
par(mfrow = c(1, 1))

# E.7
cor(imdb$lbudget, imdb$lgross) # The comment is left to the student

# E.8
cor(imdb$duration, imdb$lgross) # The comment is left to the student
```
:::
